---
title: "Clinical Trials Part 1"
author: "David Andai"
date: "10/28/2022"
output: github_document
---
A well-functioning health system ensures equitable access to essential medical products, vaccines and technologies of assured quality, safety, efficacy and cost-effectiveness, and their scientifically sound and cost-effective use. (WHO building blocks)
The noble SARS covid 19 virus, had pharmaceutical companies in a foot race to discover a vaccine that was safe for everyone. The efficacy of the vaccine and it being safe to use across all populations became a question that arose. With the global north leading in vaccine production, this possed the question of where does Africa stand in the space of drug discovery and clinical trials. 

(image)

Clinical Trials
Clinical trials are conducted to collect data regarding the safety and efficacy of new drug and device development. This paves way for drugs in manufacturing to benefit the population at hand. 
Clinical trials are an essential part of drug development. In Kenya for example, The Pharmacy and Poisons Board will approve protocols for clinical trials for new drugs and established guidelines for clinical trials involving drugs already registered in Kenya. Clinical trials are distributed into Phases.
Phases
Kenya has had the capacity to register according to the data repository at clinical trials.org
The studies are in the following phases.
* Phase I studies assess the safety of a drug or device.  usually includes a small number of healthy volunteers (20 to 100) who are generally paid for participating in the study. 
* Phase II studies test the efficacy of a drug or device. Most phase II studies are randomized trials.
* Phase III studies involve randomized and blind testing in several hundred to several thousand patients.  70% to 90% of drugs that enter Phase III studies successfully complete this phase of testing. 
* Phase IV studies, often called postmarketing surveillance trials, are conducted after a drug or device has been approved for sale. 
Pharmaceutical companies have several objectives at this stage: 
    (1) to compare a drug with other drugs already in the market
    (2) to monitor a drug's long-term effectiveness and impact on a patient's quality of life
    (3) to determine the cost-effectiveness of a drug therapy relative to other traditional and new therapies. (https://www.centerwatch.com/blog)


```{r setup, include=FALSE}
## Data Analysis on Registered clinical trials
knitr::opts_chunk$set(echo = TRUE)
#install.packages("XML")
library("XML")
library(ggplot2)
library(dbplyr)
library(tidyverse)
library(lubridate)
library(rio)
library(scales)
```

```{r include=FALSE, warning=FALSE}
#Load Clinical Trials in Kenya Data Set
setwd("C:/Users/user/Documents/R/Clinicals/Clinical_Trials")
clinicals <- xmlToDataFrame("ICTRP-Results.xml")

#Transform Date Column from character to Date variable
class(clinicals$Date_registration) ##'character'
#using the lubridate library
clinicals <- clinicals %>%
  mutate(Date_registration = lubridate::dmy(Date_registration))

class(clinicals$Date_registration)##'Date'
```

```{r include=FALSE, warning=FALSE}
# Asses The Phases of the clinical trials.
# First need to clean the data
# get the Phase column
Phase <- clinicals$Phase
#set to lower case
Phase <- tolower(Phase)
#Remove hyphens
Phase <- gsub("-/@#$:", "", Phase )
#Remove \n
Phase <- gsub("\n", "", Phase)


##Phase 4
Phase <- gsub('phase-4','phase 4', Phase)
Phase <- gsub('phase iv','phase 4', Phase)
Phase <- gsub('phase 1v','phase 4', Phase)
Phase <- gsub('iv','phase 4', Phase)

##Phase3
Phase <- gsub('phase-3','phase 3', Phase)
Phase <- gsub('phase iii','phase 3', Phase)


#Phase2
Phase <- gsub('phase-2','phase 2', Phase)
Phase <- gsub('phase ii','phase 2', Phase)
Phase <- gsub('ii','phase 2', Phase)
Phase <- gsub('phase 2i','phase 2', Phase)
Phase <- gsub('phase2/phase 2i','phase 2', Phase)
Phase <- gsub ('phase 2/phase 2','phase 2', Phase)

#Phase1
Phase <- gsub('phase-1','phase 1', Phase)
Phase <- gsub('phase i','phase 1', Phase)


#Phase0
Phase <- gsub('phase-0','phase 0', Phase)
Phase <- gsub('0','phase 0', Phase)
Phase <- gsub('phase phase 0','phase 0', Phase)

#Phase2/3
Phase <- gsub('phase 2/ phase 3','phase 2 / phase 3', Phase)
Phase <- gsub('phase 2/phase 3','phase 2 / phase 3', Phase)

#Phase1/2
Phase <- gsub('phase 1/phase 2','phase 1 / phase 2', Phase)
Phase <- gsub('phase 1/ phase 2','phase 1 / phase 2', Phase)

#Replace the long texts to empty strings
Phase <- gsub('explanatory', '', Phase)
Phase <- gsub('trials','', Phase)
Phase <- gsub('[()]', '', Phase)
Phase <- gsub('human', '', Phase)
Phase <- gsub('pharmacology', '', Phase)
Phase <- gsub('confirmatory', '', Phase)
Phase <- gsub('exploratory', '', Phase)
Phase <- gsub('therapeutic', '', Phase)
Phase <- gsub('use', '', Phase)
Phase <- gsub('no', '', Phase)
Phase <- gsub('n/a', '', Phase)

#Substitute the remaining patterns
Phase <- gsub('phase 1: yes', 'yes1', Phase)
Phase <- gsub('phase 2: yes', 'yes2', Phase)
Phase <- gsub('phase 3: yes', 'yes3', Phase)
Phase <- gsub('phase 4: yes', 'yes4', Phase)

Phase <- gsub('phase 1:', '', Phase)
Phase <- gsub('phase 2:', '', Phase)
Phase <- gsub('phase 3:', '', Phase)
Phase <- gsub('phase 4:', '', Phase)

Phase <- gsub("-","", Phase)
Phase <- gsub("[[:space:]]", "", Phase)

Phase <- gsub("yes1", "phase1", Phase)
Phase <- gsub("yes2", "phase2", Phase)
Phase <- gsub("yes3", "phase3", Phase)
Phase <- gsub("yes4", "phase4", Phase)

Phase <- gsub("/","", Phase)
Phase <- gsub("[[:space:]]", "", Phase)

Phase <- gsub('postmarketingsurveillance', 'phase4', Phase)
Phase <- gsub('tspecified','', Phase)
Phase <- gsub('tapplicable','Not_Applicable', Phase)#not_applicable

Phase <- gsub('3','phase3', Phase)
Phase <- gsub('2','phase2', Phase)

Phase <- gsub('phasephase','phase', Phase)
Phase <- gsub('phase', 'p', Phase)

clinicals["fix_Phase"] <- Phase
```

```{r include=FALSE,warning=FALSE}

#Obtain Clinical Trials from 2022
this_year <- clinicals[clinicals$Date_registration >= "2022-01-01" & clinicals$Date_registration <= "2022-09-30", ]
#delete last 3 rows, they contain NA values
this_year <- this_year[-c(74,75,76),]

#create a count column, to enable summation? 
count_a <- rep(1,times =73) #add it to this_year table
this_year <- cbind(this_year, count_a)

# Create a separate data frame, That contains This year's clinical trials and the counta column
Phases_data <- this_year[ , c("fix_Phase", "count_a")]

#Change all empty cells to NA
Phases_data [Phases_data == ""] <- NA

Phases_data <- drop_na(Phases_data)
#Then drop all Rows with NA variable
Phases_data <- na.omit(Phases_data)
Phases_data1 <- Phases_data %>% 
  group_by(fix_Phase) %>% 
  summarise_each(funs(sum))

```

```{r warning=FALSE, message=FALSE}
#Now, record each observation of phases into their own row
 plot1 <- ggplot(data = Phases_data1,
         aes(x =count_a , y =fix_Phase, label = round (count_a, 1)))+
    geom_point()+
    geom_segment(aes(y=fix_Phase, yend=fix_Phase, x=0, xend=count_a),
                 size = 1,
                 color = 'grey')+
    geom_point(
      size = 6,
      color = '#5D3FD3'
    )+
    geom_text(aes(label=count_a),color='white', size = 3.5)+
    theme_light()+
    theme (
      panel.grid.major = element_blank(),
      panel.border = element_blank(),
      axis.ticks.x = element_blank()
    )+
    xlab("")+
    ylab("Phases")+
    ggtitle("Clinical Trials Phases Distribution, Kenya")+
    labs (subtitle = '01-Jan to 30-Sep 2022')+
    scale_x_continuous(breaks = breaks_width(5))
```

```{r message=FALSE, warning=FALSE, dev='png'}
plot1
```